---
title: "Introduction to Statistical Data Analysis <br/> <br/>"
author: <small>Saad Arif <br/>  Dept. of BMS & Centre for Functional Genomics, Oxford
  Brookes University</small>
date: "<small>Nov 2019</small>"
output:
  ioslides_presentation:
    incremental: yes
  incremental: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Course Organization 



## Course Organization 



## Four Aspects of a Statistical Analysis <br /> <br /> 


1. Experimental Design
2. Exploratory Data Analysis (EDA)
3. Inference
4. Visualisation


## Experimental Design {.build}

- This is the part of statistics that happens before your experiment.

- The main aim of experimental design is to help establish a causal connection 
between the **independent** and **dependent** variable(s).

- A major goal of experimental design is to:
      + avoid  bias
      + minimize sampling error
  
- **Independent variable**: These are the **factors**/variables you control in your experiment. Sometimes also known as the predictor variables. **Dependent variable**: This is the response variable you are interested in.


## What is Bias?

- Most of the time will be working with a *sample* from a larger *population* (we can't measure the entire population for various reasons)

- We need a representative/*random* sample from the population to make valid inferences (almost all statistical tests you will ever use implicitly assume random sampling)

- **Sources of Bias**: convenience sampling, voluntary response, lack of blinding

## Avoiding Bias

- Use of control groups

- Randomization 

- Blinding

## Sampling and Sampling Error

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

